\documentclass[fontsize=11pt]{scrartcl}

%\usepackage[a5paper]{geometry}
\usepackage[a5paper, top=0.8in, bottom=0.4in, left=0.4in, right=0.6in]{geometry}

%\usepackage{fullpage}
%opening
\title{Thesis propositions accompanying the dissertation \textbf{Numerical simulations of dwarf galaxies in the Fornax Cluster}}
\author{Michele Mastropietro}

\begin{document}

%{\centering\Large\makeatletter\@title\makeatother}

  \begin{center}
%  {\centering
Thesis propositions accompanying the dissertation: \\[1ex]
    \textbf{Numerical simulations of dwarf galaxies \\ in the Fornax Cluster}

  \end{center}


\pagestyle{empty}
\begin{enumerate}
  \item High definition simulations can be a valuable tool to investigate the evolution of late-type galaxies falling into a cluster.
  \item For dwarf galaxies falling into a cluster like Fornax, the jellyfish phenomenon is a relatively short transitory phase of their evolution along the orbit, and it can be seen as a precursor of the transformation of a dwarf galaxy into an Ultra Diffuse Galaxy (UDG).
  \item We propose an estimate of the the most likely position of a galaxy with NGC 1427A-like features falling into the cluster and its most likely flight direction (Chapter~4).
  \item Simulated dwarf galaxies show an aligment of their main direction of elongation with the center of the cluster if the orbital phase lays within a quarter of the radial period after pericenter.
  \item Topological analysis with identification of low-dimensional manifolds can be used as a method to analyze the tails of a simulated jellyfish galaxy (Chapter~5).
  \item The $\lambda_R$ parameter (Emsellem et al. 2007)
  should be used with care when trying to assess apparent angular momentum of a galaxy.
  \item Software code should be written to be read by humans, not by machines. Knowledge embedded into code should be made as clear as possible to future users and readers.
  \item Code used for scientific research should be thought and designed as being public from the very first line of code (regardless of the time of its actual publication). Doing so encourages to write code for reproducible science. Open source software has been key to fly a helicopter on Mars!
%  \item The modern approach to software development and engineering (e.g. use of collaborative software platforms, version control, unit testing) should be taught from bachelor programs.

%  \item Software for simulation of the galaxy formation suffered a large fragmentation of efforts. Many of these efforts were carried out by people not interested in reusability because of their contract terms. Supervisors should take care of this. This hinders reproducibility.

%  \item Use of collaborative software platforms should be encoraged. No need to build the wheel again and again, only to have it in a different colour :P.
%  \item mental health
%  \item For late-type dwarf galaxies as the one we simulated, $\lambda_R$ is affected by...

%  \item cura persone e ambiente:, scrivere bene il software

\end{enumerate}

\begin{flushright}
  Michele Mastropietro
  \end{flushright}

\end{document}
